Manuscript: Detecting differences in Size Spectra

### 2023-01-11 Manuscript.Rmd
new version knit for coauthors jan 11 2023

Body size distributions are a fundamental characteristic of communities. In general, size-abundance relationships are expected to scale with the value of -0.75 across all biological communities (Damuth 1981, 1991, 1998). This is thought to be a consequence of simple size-dependent metabolic constraints on organisms’ energy use predicted by the metabolic theory of ecology (Nee 1991, Brown et al. 2004). The remarkable consistency of these relationships across spatiotemporal scales and ecosystems has led them to be recommended as a "universal" indicator of ecological status (Petchey and Belgrano 2010)[@Petchey2010]. Variation in size-abundance relationships have been documented through space (Pomeranz 2021), time (newish Sprules papers?), seasonality (McGarvey and Kirk), to human activities (fish abundance with no fishing paper). Likwise, variation in size spectra relationships have been used to explain fundamental differences in how communities are organized. For example, external resource subsidies "bend the rules" and allow higher abundances of large body sizes than would be expected (Perkins). However, recent research has shown that these results may be an artifact of how the data were treated. Edwards et al. (2020?) reanalyzed a time series of marine fisheries data and found that previously reported changes through time were actually dependent on the methodolugy used.

Individual size distributions (ISD *sensu* [@White2007]), also referred to as abundance size spectra) are one of the size-abundance relationships commonly used. ISDs show the relationship between the abundances of individual body size regardless of taxonomy. Generally, there is a negative relationship between individual body size (M) on the *x*-axis  and abundance (N) on the *y*-axis. Theoretical and empirical data support this relationship being described as a simple power law with exponent $\lambda$ in the form of $N \sim M^{\lambda}$, where $\lambda = 2$. Commonly, $N$ is the count of body sizes grouped into bins, and $\lambda$ is estimated as the slope from OLS regressions of log transformed data of $N$ and the mid point of the body size bin $M_{bin}$: $\log(N_{count}) = \lambda \log(M_{bin})$. Myriad binning methods have been proposed, including linear and logarithmic bin widths. Likewise, methods can rely on the absolute count in the bins as well as normalization techniques where the count is divided by the bin width (especially common with logarithmic binning). Alternatively, $\lambda$ can be estimated directly using maximum likelihood techniques.

Previous work has shown that the estimates of $\lambda$ differ between MLE and OLS techniques. OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2008, Edwards et al. 2017), and reanalysis of empirical data also indicates that the conclusions are dependent on the methodology used [@white2008 @edwards2020]. However, recent empirical analysis of stream macroinvertebrate communities across the National Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied, the relative change across the environmental gradient was consistent regardless of method used [@Pomeranz2022]. While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of OLS slope parameters of 0.1 coincide with a relative change of MLE estimates of 0.1?

@@

In order to answer the first two questions, we simulate body size observations from bounded power law distributions with varied $\lambda$ exponents across a hypothetical environmental gradient and compare the results obtained using three different methodologies common in the literature. To answer the third question, we re-analyze two previously published data sets of stream community body sizes across a stress and environmental gradient, respectively. We find that the MLE method more accurately estimates the site-specific $\lambda$ exponents as well as the relative change in exponents across the hypothetical gradient, as well as having smaller variation around the estimates. The logarithmic binning methods generally perform well, but have larger variation around the site-specific estimates. In addition, the variation of the estimates of the hypothetical relationship obtained from the binning methods is dependent on the value of $\lambda$ used (under and over estimates steep and shallow $\lambda$, respectively).

@@

In order to investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in [@Edwards2017]. Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values are based on empirical body sizes of stream benthic communities reported in [@Pomeranz2022]. Our results are not dependent on the range of body sizes, and we show the results of other body size ranges in the supplemental information.

@@

For the first binning method, we created 6 equal logarithmic bins covering the range of body sizes. This method has been used extensively in previous studies. For example see [@Perkins2018, @martinez2016, @dossenaWarmingAltersCommunity2012]. In the present study, the count in each bin was normalized by dividing by the bin width to account for the unequal bin sizes. The process of normalizing shifts the estimate by -1. In other words, an un-normalized estimate of -0.75 would result in an estimate of -1.75 when normalizing the data [@sprulesSurfingBiomassSize2015, Edwards2018, Pomeranz2022]. Throughout the manuscript, the normalized equal logarithmic binning method will be referred to as ELBn.

### Log~2~ Bins: NAS
The second method was similar to ELBn, but bins of Log~2~ widths are used. The number of bins is dependent on the size ranges present in the data. When working with empirical data with different size ranges this can alter the number of bins per site. However, since the data here were simulated from a known size range, the number of bins for each site is identical. The count in each bin is normalized in the same way. This method has been  used [@jennings2004, @Jennings2002, @sprulesSurfingBiomassSize2015, @pomeranzAnthropogenicMiningAlters2019; @mcgarveySeasonalComparisonCommunitylevel2018) and is referred to as the Normalized Abundance Spectrum (NAS).

For the two binning methods, simple OLS regression were conducted in the form $Log_{10}(N) = \beta_0 + \beta_1 * Log_{10}(M)$. Where $\beta_{1}$ is considered to be the $\lambda_{estimate}$.

@@

Previous work has investigated the biases in $\lambda$ estimates when using different methods [-@Edwards2017, @edwards2020, @White2007]. However, the focus of the present work is to investigate biases when estimating the relationship parameters across an environmental gradient.

@@

Each sample of body sizes from different $\lambda$ values was assumed to come from a community across a hypothetical environmental gradient $X$. Values of $X$ for each community are uniformly distributed from -1 to 1 (site-specific x-values: $(-1.0, -0.5, 0, 0.5, 1.0)$. For each $\lambda$ "window", the relative change in $\lambda$ values ($\beta_{env}$) across the gradient is equal to -0.5, such that our model is:

@@

We re-analyze two data sets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage (AMD) stress gradient. Details of the sample collection and processing can be found in [@pomeranzAnthropogenicMiningAlters2019]. Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using taxon-specific published length weight regressions.

The second data set was from the wadeable stream sites of the National Ecological Observatory Network [@NEON_Inverts2022]. NEON stream sites are located across a wide temperature gradient in the United States, from Puerto Rico to Alaska. Quantitative macroinvertebrate samples were collected using the most appropriate method based on the local habitat. All individuals were identified and had their body lengths measured, and body mass was estimated using published length weight regressions. This data has been analyzed previously using size spectra methods as described in [@Pomeranz2022]. Detailed methods of the sampling collection and data processing methods can be found on the NEON website ([<https://www.neonscience.org/data-collection/macroinvertebrates>], or macroinvertebrate DPI pubs). Coefficient estimates ($\pm$ 1 standard deviation) are compared across methods. This allows us to determine whether or not the main results would differ depending on the method used.

@@

The data used in this manuscript is already publicly available [@Pomeranz2019] data dryad DOI <https://datadryad.org/stash/dataset/doi:10.5061/dryad.v6g985s>, [@NEON_Inverts2022]. R Scripts to reproduce the full simulation and analysis are available at: ([<https://github.com/Jpomz/detecting-spectra-differences>]).

@@

![Figure 6. of empirical data estimates. All of the methods estimate the same sign of the relationship, but the estimates from the binning methods are generally greater than the MLE estimates. **NOTE** I need to re-plot these in ggplot and and stitch together. Currently stitching pre-made png files, I think that's why it looks so funky.](figures/empirical_combined.png)

@@

The relationship between body size and abundance has been extensively studied in a wide range of taxa inhabiting both terrestrial and aquatic ecosystems (reviewed by [@Brown1995, GastonBlackburn2000, @White2007]). Empirical patterns are remarkably consistent, and can be explained by the metabolic theory of Ecology [@Brown2004].  Measuring parameters describing the decline in abundance with increasing body size in communities is being done with increasing frequency across ecology. Previous work has investigated the accuracy and inherent biases associated with different estimation methods. However, how these inaccuracies and biases compound across environmental gradients remains uncertain, making it difficult to detect variation in size spectra parameters across environmental gradients with confidence. Here, we sampled body sizes from known distributions with varying parameters ($\lambda$) and estimated the coefficient of the relationship ($\beta_{env}$) across a hypothetical gradient in order to assess how the results would vary depending on the method used. Likewise, we compared how the interpretation of previously published results could change depending on the methodology used to estimate the relationship.

@@

The MLE method outperformed both of the binning methods under nearly any measure. With the publication of the *sizespectra* package [-@Edwards2017], producing MLE estimates of size spectra parameters is a relatively easy task. Therefore, we recommend using this in all future studies of size spectra relationships.

We reiterate the recommendations of [-@White2007, -@sprulesSurfingBiomassSize2015, -@Edwards2017] to estimate size spectra relationships using MLE methods due to their superior performance in nearly every context. Furthermore, we strongly encourage authors to publish individual size data whenever possible. This will allow for the consistent re-analysis of existing data sets as methodologies develop and improve. This will aid in the ability for size spectra work to be synthesized between research groups and across scales.

@@

>>>>

Body size distributions are a fundamental characteristic of communities. In general, size-abundance relationships are expected to scale with the value of -0.75 across all biological communities (Damuth 1981, 1991, 1998). This is thought to be a consequence of simple size-dependent metabolic constraints on organisms’ energy use predicted by the metabolic theory of ecology (Nee 1991, Brown et al. 2004). The remarkable consistency of these relationships across spatiotemporal scales and ecosystems has led them to be recommended as a "universal" indicator of ecological status [@Petchey2010]. Variation in size-abundance relationships have been documented through space [@pomeranz2022], time [@evans2022], seasonality [@mcgarvey2018], to human activities [@jennings2004]. Likewise, variation in size spectra relationships have been used to explain fundamental differences in how communities are organized. For example, external resource subsidies "bend the rules" and allow higher abundances of large body sizes than would be expected [@perkins2018]. However, recent research has shown that these results may be an artifact of how the data were treated. Edwards et al. (2020) reanalyzed a time series of marine fisheries data and found that previously reported changes through time were actually dependent on the methodology used.

Individual size distributions (ISD *sensu* [@White2007], also referred to as abundance size spectra) are one of the size-abundance relationships commonly used. ISDs show the relationship between the abundances of individual body size regardless of taxonomy. Generally, there is a negative relationship between individual body size (M) on the *x*-axis  and abundance (N) on the *y*-axis. Theoretical and empirical data support this relationship being described as a simple power law with exponent $\lambda$ in the form of $N \sim M^{\lambda}$, where $\lambda = 2$ [@Sheldon1972, @Andersen2006a]. Commonly, $N$ is the count of body sizes grouped into bins, and $\lambda$ is estimated as the slope from OLS regressions of log transformed data of $N$ and the mid point of the body size bin $M_{bin}$: $\log(N_{count}) = \lambda \log(M_{bin})$. Myriad binning methods have been proposed, including linear and logarithmic bin widths. Likewise, methods can rely on the absolute count in the bins as well as normalization techniques where the count is divided by the bin width (especially common with logarithmic binning). Alternatively, $\lambda$ can be estimated directly using maximum likelihood techniques.

Previous work has shown that the estimates of $\lambda$ differ between MLE and OLS techniques. OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2008, Edwards et al. 2017), and reanalysis of empirical data also indicates that the conclusions are dependent on the methodology used [@white2008 @edwards2020]. However, recent empirical analysis of stream macroinvertebrate communities across the National Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied based on method, the relative change across the environmental gradient was consistent across methods [@pomeranz2022]. While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of OLS slope parameters of 0.1 coincide with a relative change of MLE $\lambda$ estimates of 0.1?

@@

In order to answer the first two questions, we simulate body size observations from bounded power law distributions with varied $\lambda$ exponents across a hypothetical environmental gradient and compare the results obtained using three different methodologies common in the literature. To answer the third question, we re-analyze two previously published data sets of community body sizes across a stress and environmental gradient, respectively [@pomeranz2019, @pomeranz2022]. We find that the MLE method more accurately estimates the site-specific $\lambda$ exponents as well as the relative change in exponents across the hypothetical gradient, as well as having smaller variation around the estimates. The logarithmic binning methods generally perform well, but have larger variation around the site-specific estimates. In addition, the variation of the estimates of the hypothetical relationship obtained from the binning methods is dependent on the value of $\lambda$ used (under and over estimates steep and shallow $\lambda$, respectively).

@@

In order to investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in [@edwards2017]. Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values are based on empirical body sizes of stream benthic communities reported in [@pomeranz2022]. Our results are not dependent on the range of body sizes, and we show the results of other body size ranges in the supplemental information.

@@

For the first binning method, we created 6 equal logarithmic bins covering the range of body sizes. This method has been used extensively in previous studies. For example see [@perkins2018, @martinez2016, @dossena2012]. In the present study, the count in each bin was normalized by dividing by the bin width to account for the unequal bin sizes. The process of normalizing shifts the estimate by -1. In other words, an un-normalized estimate of -0.75 would result in an estimate of -1.75 when normalizing the data [@sprules2015, @edwards2017, @pomeranz2022]. Throughout the manuscript, the normalized equal logarithmic binning method will be referred to as ELBn.

### Log~2~ Bins: NAS
The second method was similar to ELBn, but bins of Log~2~ widths are used. The number of bins is dependent on the range of body sizes present in the data. When working with empirical data with different size ranges this can alter the number of bins per site, and it is recommended to construct the bins based on the global min and max body size present. However, since the data here were simulated from a known size range, the number of bins for each site is identical. The count in each bin is normalized in the same way as described above. This method has been  used [@jennings2004, @Jennings2002, @sprules2015, @pomeranz2019; @mcgarvey2018) and is referred to as the Normalized Abundance Spectrum (NAS).

For the two binning methods, simple OLS regression were conducted in the form 
$$Log_{10}(N) = \beta_0 + \beta_1 * Log_{10}(M)$$

Where $\beta_{1}$ is considered to be the $\lambda_{estimate}$.

@@

Previous work has investigated the biases in $\lambda$ estimates when using different methods [-@edwards2017, @edwards2020, @White2007]. However, the focus of the present work is to investigate biases when estimating the relationship parameters across an environmental gradient.

@@

Each sample of body sizes from different $\lambda$ values was assumed to come from a community across a hypothetical environmental gradient $X$. Values of $X$ for each community are uniformly distributed from -1 to 1 (site-specific x-values: $(-1.0,~-0.5,~0,~0.5,~1.0)$. For each $\lambda$ "window", the relative change in $\lambda$ values ($\beta_{env}$) across the gradient is equal to -0.5, such that our model is:

@@

We re-analyze two data sets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage (AMD) stress gradient. Details of the sample collection and processing can be found in [@pomeranz2019]. Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using taxon-specific published length weight regressions.

The second data set was from the wadeable stream sites of NEON [@NEON_Inverts2022]. NEON stream sites are located across a wide temperature gradient in the United States, from Puerto Rico to Alaska. Quantitative macroinvertebrate samples were collected using the most appropriate method based on the local habitat. All individuals were identified and had their body lengths measured, and body mass was estimated using published length weight regressions. This data has been analyzed previously using size spectra methods as described in [@pomeranz2022]. Detailed methods of the sampling collection and data processing methods can be in the macroinvertebrate DPI pubs found on the NEON website <https://www.neonscience.org/data-collection/macroinvertebrates>. Coefficient estimates ($\pm$ 1 standard deviation) are compared across methods. This allows us to determine whether or not the main results would differ depending on the method used.

@@

The data used in this manuscript is already publicly available [@Pomeranz2019] data dryad DOI <https://datadryad.org/stash/dataset/doi:10.5061/dryad.v6g985s>, [@NEON_Inverts2022]. R Scripts to reproduce the full simulation and analysis are available at: ([<https://github.com/Jpomz/detecting-spectra-differences>]). (To be archived upon acceptance.)

@@

![Figure 6. of empirical data estimates. All of the methods estimate the same sign of the relationship, but the estimates from the binning methods are generally greater than the MLE estimates.](figures/empirical_combined.png)

@@

The relationship between body size and abundance has been extensively studied in a wide range of taxa inhabiting both terrestrial and aquatic ecosystems (reviewed by [@brown1995, GastonBlackburn2000, @White2007]). Empirical patterns are remarkably consistent, and can be explained by the metabolic theory of Ecology [@Brown2004].  Measuring parameters describing the decline in abundance with increasing body size in communities is being done with increasing frequency across ecology. Previous work has investigated the accuracy and inherent biases associated with different estimation methods. However, how these inaccuracies and biases compound across environmental gradients remains uncertain, making it difficult to detect variation in size spectra parameters across environmental gradients with confidence. Here, we sampled body sizes from known distributions with varying parameters ($\lambda$) and estimated the coefficient of the relationship ($\beta_{env}$) across a hypothetical gradient in order to assess how the results would vary depending on the method used. Likewise, we compared how the interpretation of previously published results could change depending on the methodology used to estimate the relationship.

@@

The MLE method outperformed both of the binning methods under nearly any measure. With the publication of the *sizespectra* package [-@edwards2017], producing MLE estimates of size spectra parameters is a relatively easy task. Therefore, we recommend using this in all future studies of size spectra relationships.

We reiterate the recommendations of [-@White2007, -@sprules2015, -@edwards2017] to estimate size spectra relationships using MLE methods due to their superior performance in nearly every context. Furthermore, we strongly encourage authors to publish individual size data whenever possible. This will allow for the consistent re-analysis of existing data sets as methodologies develop and improve. This will aid in the ability for size spectra work to be synthesized between research groups and across scales.

@@

```{r, echo=TRUE}
# kable() above looks awful
# here is another bad option
# if we decide to include this I'll make it look nice in excel. 
read.csv("lit comparison.csv")
```
